---
title: "IncuCyte Confluency Report"
subtitle: "Per-plate summary and heatmaps"
date: today
format:
  html:
    toc: true
    embed-resources: true
    df-print: paged
    code-fold: true
    code-summary: "Show code"
---

```{css sticky-header, echo=FALSE}
.table-scroll-sticky thead th {
  position: sticky;
  top: 0;
  z-index: 2;
  background-color: #CCCCCC;
}
```

```{r setup}
library(tidyverse)
library(knitr)
library(kableExtra)
library(here)

# --- Data location ---------------------------------------------------------
# Root folder to search for IncuCyte export files (see `depth` in Parse data,
# below, for how far under this folder the search recurses).
data_dir <- here("..")

# --- parse_confluency.R -----------------------------------------------------
# Caches a local copy on first run so the report doesn't re-fetch on every 
# render; delete to force a re-download of the latest version.
script_path <- here("R", "parse_confluency.R")
script_url  <- "https://raw.githubusercontent.com/jackspepper/tools/main/incucyte_parse_confluency/parse_confluency.R"

if (!file.exists(script_path)) {
  dir.create(here("R"), showWarnings = FALSE)
  download.file(script_url, script_path, mode = "wb")
}
source(script_path, echo = FALSE)

# --- Output ------------------------------------------------------------------
out_dir <- here("output")
dir.create(out_dir, showWarnings = FALSE)

# Standard styled + scrollable kable, with a pinned header row.
# height: max box height before scrolling kicks in.
scrolly_kable <- function(tbl, height = "500px", ...) {
  kable(tbl, ...) |>
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE,
      position = "center",
      font_size = 14
    ) |>
    row_spec(0, bold = TRUE, background = "#CCCCCC") |>
    scroll_box(height = height, width = "100%") |>
    gsub('class="table table-striped', 'class="table-scroll-sticky table table-striped', x = _, fixed = TRUE)
}
```

This report parses IncuCyte confluency exports into a tidy table, then summarises, flags, and visualises well-level confluence. Adjust **Study configuration** and **Exclusions** below to fit a given run — the rest of the report updates automatically.

## Parse data

```{r parse}
result <- read_confluency_folder(
  data_dir,
  depth = 5,
  pattern = "Plate.*\\.txt$",
  verbose = TRUE,
  progress = TRUE,
  export = TRUE
)

data <- result$data |>
  mutate(
    date          = str_extract(file_path, "\\d{8}(?=_)"),
    plate         = str_extract(file_path, "Plate\\d+"),
    imaging_date  = str_extract(file_path, "(?<=_Imaged)\\d{2}[A-Z]{3}\\d{4}"),
    imaging_type  = str_extract(file_path, "(?<=_)[^_]+(?=\\.txt$)")
  )

well_digits <- max(nchar(as.character(data$column)))
data <- data |>
  mutate(well = paste0(row, formatC(column, width = well_digits, flag = "0")))

write_confluency_output(result, out_dir = out_dir)
```

`imaging_date` and `imaging_type` are parsed from the folder segment following the plate date, e.g. `20260720_Imaged27JUL2026_Prefix` → imaging date `20JUL2026`, type `Prefix`. Plates sharing the same initial `date` are the same set; `imaging_type` distinguishes imaging rounds within that set. Both are `NA` where the pattern isn't present in the path, and rows are kept regardless.

Rows and columns below auto-detect from the data (no hardcoded 8×12/96-well assumption), so 6-, 24-, 96-, or 384-well plates all work without editing the report. If a plate is mid-imaging or missing wells, its row/column range in the heatmaps reflects whatever is present, not the full nominal plate.

## Study configuration

Settings that vary by study — edit these, then run the rest of the report unchanged.

```{r stage-config}
# Assay stages present in this study, earliest -> latest. Drives the stage
# comparison section below; stage names not listed here are excluded from
# that section only (they still appear in the summary/heatmaps above).
# e.g. c("Baseline", "Endpoint") or c("Prefix", "Postfix", "Preassay", "Postassay")
#
# A single "..." entry can be used in place of (or in addition to) explicit
# stage names. It expands to any imaging_type values found in the data that
# aren't already listed elsewhere in stage_levels, inserted alphabetically at
# that position. This lets you pin specific stages (e.g. a known first/last)
# while auto-including whatever else shows up.
# e.g. c("Prefix", "...", "Preassay") or c("...") or c("Baseline", "...")
stage_levels <- c("Prefix", "Postfix", "...", "Preassay")

if ("..." %in% stage_levels) {
  dots_pos <- which(stage_levels == "...")[1]
  named_stages <- setdiff(stage_levels[stage_levels != "..."], NA)
  other_stages <- sort(setdiff(unique(data$imaging_type), c(named_stages, NA)))
  stage_levels <- append(
    stage_levels[stage_levels != "..."],
    other_stages,
    after = dots_pos - 1
  )
}

# Metric the stage comparison section focuses on. Must match a value in the
# `metric` column exactly (see Per-plate summary for available values).
# Set to NA to include all metrics in the comparison section instead.
primary_metric <- "Phase Object Confluence (%)"

# Acceptable confluence range, used for well flagging and the heatmap colour
# scales below. Adjust to your assay's expected range.
range_lower <- 80
range_upper <- 100
```

## Exclusions

Set any of these vectors to exclude specific dates, plates, metrics, or wells from the entire report (summary, flags, and both heatmaps). Leave a vector as `character(0)` to exclude nothing for that field.

```{r exclusions}
exclude_dates         <- character(0)  # e.g. c("20260712")
exclude_plates        <- character(0)  # e.g. c("Plate3")
exclude_metrics       <- character(0)  # e.g. c("Phase Object Confluence (%)")
exclude_wells         <- character(0)  # e.g. c("A01", "H12")
exclude_imaging_dates <- character(0)  # e.g. c("20JUL2026")
exclude_imaging_types <- character(0)  # e.g. c("Preassay")

data <- data |>
  filter(
    !date          %in% exclude_dates,
    !plate         %in% exclude_plates,
    !metric        %in% exclude_metrics,
    !well          %in% exclude_wells,
    !imaging_date  %in% exclude_imaging_dates,
    !imaging_type  %in% exclude_imaging_types
  )
```

## Summaries

### Per-date-group summary

Standard descriptive stats per date and metric.

```{r date-group summary}
summary_tbl <- data |>
  group_by(date, imaging_date, imaging_type, metric) |>
  summarise(
    n      = n(),
    mean   = mean(value, na.rm = TRUE),
    sd     = sd(value, na.rm = TRUE),
    min    = min(value, na.rm = TRUE),
    max    = max(value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(date, imaging_date, imaging_type, metric)

scrolly_kable(summary_tbl, digits = 2)
```

### Per-plate summary

Standard descriptive stats per plate, date, and metric.

```{r plate summary}
summary_tbl <- data |>
  group_by(date, plate, imaging_date, imaging_type, metric) |>
  summarise(
    n      = n(),
    mean   = mean(value, na.rm = TRUE),
    sd     = sd(value, na.rm = TRUE),
    min    = min(value, na.rm = TRUE),
    max    = max(value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(date, plate, imaging_date, imaging_type, metric)

scrolly_kable(summary_tbl, digits = 2)
```

## Flagged wells

Wells outside the expected confluence range are flagged below. Range is set in Study configuration (`range_lower` / `range_upper`).

```{r flag-range}
data <- data |>
  mutate(flagged = value < range_lower | value > range_upper)

flagged_tbl <- data |>
  filter(flagged) |>
  select(date, plate, imaging_date, imaging_type, well, metric, value) |>
  arrange(date, plate, imaging_date, imaging_type, well)
```

```{r flag-table}
if (nrow(flagged_tbl) == 0) {
  cat("No wells fell outside the range", range_lower, "-", range_upper, "%.\n")
} else {
  scrolly_kable(flagged_tbl, digits = 2, caption = paste0("Wells outside ", range_lower, "-", range_upper, "%"))
}
```

## Plate heatmaps

Well-level confluence by plate, date, and imaging type. Colour scale is fixed to `r range_lower`–`r range_upper`% (squished, from Study configuration) — wells outside this range are clipped to the scale's extremes rather than stretching it.

```{r heatmap, fig.width=12, fig.height=6}
plot_columns <- sort(unique(data$column))

date_type_levels <- data |>
  distinct(date, imaging_type) |>
  mutate(
    stage_rank = match(imaging_type, stage_levels),
    stage_rank = ifelse(is.na(stage_rank), length(stage_levels) + 1, stage_rank),
    date_type = paste0(date, ifelse(is.na(imaging_type), "", paste0(" (", imaging_type, ")")))
  ) |>
  arrange(date, stage_rank, imaging_type) |>
  pull(date_type) |>
  unique()

plot <- data |>
  mutate(
    row      = factor(row, levels = rev(sort(unique(row)))),  # A at top
    date_type = paste0(date, ifelse(is.na(imaging_type), "", paste0(" (", imaging_type, ")"))),
    date_type = factor(date_type, levels = date_type_levels)
  ) |>
  ggplot(aes(x = column, y = row, fill = value)) +
  geom_tile(color = "white") +
  coord_equal() +
  scale_x_continuous(breaks = plot_columns) +
  facet_grid(date_type ~ plate, labeller = labeller(date_type = label_wrap_gen(width = 8))) +
  labs(
    x = "Column",
    y = "Row",
    fill = "Confluence (%)"
  ) +
  theme_minimal() +
  theme(strip.text.x = element_text(size = 9)) +
  scale_fill_viridis_c(
    limits = c(range_lower, range_upper),
    oob = scales::squish
  )

plot

ggsave(file.path(out_dir, "PlateConfluencyHeatmap.png"), plot, width = 12, height = 6, units = "cm", dpi = 600, scale = 4)
```

## Flagged wells heatmap

Same layout as above, but flagged wells (outside `r range_lower`-`r range_upper`%) are highlighted in red; all other wells are shown in grey for context.

```{r flag-heatmap, fig.width=12, fig.height=6}
flag_plot <- data |>
  mutate(
    row      = factor(row, levels = rev(sort(unique(row)))),
    date_type = paste0(date, ifelse(is.na(imaging_type), "", paste0(" (", imaging_type, ")"))),
    date_type = factor(date_type, levels = date_type_levels)
  ) |>
  ggplot(aes(x = column, y = row, fill = flagged)) +
  geom_tile(color = "white") +
  coord_equal() +
  scale_x_continuous(breaks = plot_columns) +
  facet_grid(date_type ~ plate, labeller = labeller(date_type = label_wrap_gen(width = 8))) +
  labs(
    x = "Column",
    y = "Row",
    fill = paste0("Outside ", range_lower, "-", range_upper, "%")
  ) +
  theme_minimal() +
  theme(strip.text.x = element_text(size = 9)) +
  scale_fill_manual(values = c(`FALSE` = "grey85", `TRUE` = "red"))

flag_plot

ggsave(file.path(out_dir, "PlateFlaggedWellsHeatmap.png"), flag_plot, width = 12, height = 6, units = "cm", dpi = 600, scale = 4)
```

## Assay stage comparison

Compares confluence across the assay stages defined in `stage_levels` (Study configuration, above) to track how each well changes between prep and use. Only stage pairs actually present for a given well/date/plate/metric are compared — missing images are skipped rather than treated as zero or NA-flagged. Any `imaging_type` values not listed in `stage_levels` are excluded from this section only. Focused on `r if (is.na(primary_metric)) "all metrics" else primary_metric` (set `primary_metric` in Study configuration).

```{r stage-compare-wide}
stage_input <- if (is.na(primary_metric)) {
  data
} else {
  data |> filter(metric == primary_metric)
}

stage_wide <- stage_input |>
  filter(imaging_type %in% stage_levels) |>
  mutate(imaging_type = factor(imaging_type, levels = stage_levels)) |>
  select(date, plate, well, row, column, metric, imaging_type, value) |>
  pivot_wider(names_from = imaging_type, values_from = value)

# Ensure a column exists for every configured stage, even if absent from the dataset
for (col in stage_levels) {
  if (!col %in% names(stage_wide)) stage_wide[[col]] <- NA_real_
}

# Every consecutive pair (adjacent) plus every stage vs. the first (baseline), skipping duplicates
consecutive_pairs <- if (length(stage_levels) >= 2) {
  purrr::map2(stage_levels[-length(stage_levels)], stage_levels[-1], c)
} else list()

baseline_pairs <- if (length(stage_levels) >= 3) {
  purrr::map(stage_levels[-1], ~ c(stage_levels[1], .x))
} else list()

stage_pairs <- unique(c(consecutive_pairs, baseline_pairs))
stage_pairs <- Filter(function(p) p[1] != p[2], stage_pairs)

delta_names <- purrr::map_chr(stage_pairs, ~ paste(.x[2], "-", .x[1]))

comparison_lookup <- tibble(
  comparison = delta_names,
  from_stage = purrr::map_chr(stage_pairs, 1),
  to_stage   = purrr::map_chr(stage_pairs, 2)
)

stage_deltas <- stage_wide
for (i in seq_along(stage_pairs)) {
  from_stage <- stage_pairs[[i]][1]
  to_stage   <- stage_pairs[[i]][2]
  stage_deltas[[delta_names[i]]] <- stage_deltas[[to_stage]] - stage_deltas[[from_stage]]
}
```

### Per-well changes

```{r stage-compare-well-table}
well_delta_tbl <- stage_deltas |>
  select(date, plate, well, metric, all_of(stage_levels), all_of(delta_names)) |>
  filter(rowSums(!is.na(across(c(Prefix, Postfix, Preassay)))) >= 2) |>
  arrange(date, plate, well, metric)

scrolly_kable(well_delta_tbl, digits = 2)
```

### Per-plate summary of changes

Mean/SD of each stage-to-stage delta across wells, computed only from wells where both stages in that pair are present.

```{r stage-compare-plate-summary}
delta_long <- stage_deltas |>
  select(date, plate, well, metric, all_of(stage_levels), all_of(delta_names)) |>
  pivot_longer(
    cols = all_of(delta_names),
    names_to = "comparison",
    values_to = "delta"
  ) |>
  filter(!is.na(delta)) |>
  left_join(comparison_lookup, by = "comparison") |>
  rowwise() |>
  mutate(
    from_value = get(from_stage),
    to_value   = get(to_stage)
  ) |>
  ungroup()

plate_delta_summary <- delta_long |>
  group_by(date, plate, metric, comparison) |>
  summarise(
    n_wells = n(),

    mean_from = mean(from_value),
    mean_to   = mean(to_value),

    mean_delta = mean(delta),
    sd_delta   = sd(delta),
    min_delta  = min(delta),
    max_delta  = max(delta),

    .groups = "drop"
  ) |>
  mutate(
    reference = sprintf(
      "%.1f → %.1f (Δ %.1f)",
      mean_from,
      mean_to,
      mean_delta
    )
  ) |>
  arrange(date, plate, metric, comparison)

scrolly_kable(plate_delta_summary, digits = 2)
```

### Delta heatmaps

Well-level change between consecutive stages present in the data, in the order defined by `stage_levels`. Diverging scale centred on 0; grey tiles indicate a pair where one or both stages are missing for that well.

```{r stage-compare-heatmap, fig.width=12, fig.height=6}
delta_plot_data <- delta_long |>
  filter(comparison %in% delta_names) |>
  mutate(comparison = factor(comparison, levels = delta_names)) |>
  mutate(row = str_extract(well, "^[A-Z]"),
         column = as.integer(str_extract(well, "\\d+$")),
         row = factor(row, levels = rev(sort(unique(row)))))

delta_plot <- delta_plot_data |>
  mutate(
    plate_label = paste0(date, " - ", comparison)
  ) |>
  ggplot(aes(x = column, y = row, fill = delta)) +
  geom_tile(color = "white") +
  coord_equal() +
  scale_x_continuous(breaks = plot_columns) +
  facet_grid(plate_label ~ plate, labeller = labeller(plate_label = label_wrap_gen(width = 30))) +
  labs(x = "Column", y = "Row", fill = "\u0394 Confluence (%)") +
  theme_minimal() +
  theme(strip.text.x = element_text(size = 9)) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, na.value = "grey85")

delta_plot

ggsave(file.path(out_dir, "PlateStageDeltaHeatmap.png"), delta_plot, width = 12, height = 6, units = "cm", dpi = 600, scale = 4)
```

## Metadata

Parsed metadata key-value pairs per source file.

```{r metadata}
scrolly_kable(result$metadata)
```
