---
title: "IncuCyte Confluency Report — Mini"
subtitle: "Stripped-down view: plate heatmaps only"
date: today
format:
  html:
    toc: true
    embed-resources: true
    df-print: paged
    code-fold: true
    code-summary: "Show code"
---

```{r read-chunks, include=FALSE}
knitr::read_chunk("IncuCyte_Confluency_Report.qmd")
```

This is a stripped-down view of `IncuCyte_Confluency_Report.qmd`, re-using the same setup, parsing, and configuration chunks but showing only the sections included below. Edit **Study configuration** and **Exclusions** here exactly as in the full report — the underlying code is shared, not duplicated.

To include or drop a section, add or remove its chunk (matching the same `{r label}` from the main report) below. Any chunk in the main report can be pulled in this way, in any order, as long as the chunks it depends on run first.

## Setup

```{r setup}
```

## Parse data

```{r parse}
```

## Study configuration

```{r stage-config}
```

## Exclusions

```{r exclusions}
```

```{r flag-range}
```

## Plate heatmaps

```{r heatmap, fig.width=12, fig.height=6}
```

## Flagged wells heatmap

```{r flag-heatmap, fig.width=12, fig.height=6}
```

## Assay stage comparison

```{r stage-compare-wide}
```

```{r stage-compare-plate-summary}
```

```{r stage-compare-heatmap, fig.width=12, fig.height=6}
```