---
title: "Running Consolidation and Understanding Outputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running Consolidation and Understanding Outputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette covers `qpcr_consolidation.R` (Script 2), the resulting output files, and the audit log. Run `vignette("cleaning-pipeline", package = "qpcrpipeline")` first. For a full list of default values for every variable mentioned below, see `vignette("reference", package = "qpcrpipeline")`.

## Running the consolidation script

After the cleaning pipeline has finished, run the consolidation script (`qpcr_consolidation.R`). This reads all the cleaned CSV files from the `outputs` folder and assembles them into two Excel workbooks.

### Step 1 - Check your settings (Section 1 of the script)

**Input directory**

```{r}
input_dir <- "outputs/"
```

This should match the `output_dir` from the cleaning pipeline.

**Output location**

```{r}
CONSOLIDATION_DIR <- "consolidated"
ALL_OUT_PATH    <- file.path(CONSOLIDATION_DIR, "qpcr_all_samples.xlsx")
REVIEW_OUT_PATH <- file.path(CONSOLIDATION_DIR, "qpcr_review_samples.xlsx")
```

**Target alias mapping**

```{r}
TARGET_ALIASES <- list(
  uni = c("uni", "universal", "univ")
)
```

Treats `uni`, `universal`, and `univ` as the same target and consolidates them onto a single sheet named `uni`. You do not normally need to change this.

### Step 2 - Run the script

Press **Ctrl+Shift+Enter** (or **Cmd+Shift+Enter** on Mac).

### The output file conflict prompt

If the output Excel files already exist from a previous run:

```
------------------------------------------------------------
 All-samples workbook already exists:
  Modified : 2024-12-10 14:32:05
  Size     : 243.1 KB
  Sheets   : run_summary, fucp, hpd3, lyta, uni (5 sheets)
------------------------------------------------------------
 Options:
   O = Overwrite  - discard existing workbook and rebuild
   A = Append     - merge new data into existing workbook
   N = New file   - save under a different name
 Enter O, A, or N: _
```

**O - Overwrite:** The existing workbook is deleted and rebuilt from scratch. Use this when you have reprocessed all plates or want a completely fresh consolidation.

**A - Append:** New data from freshly processed plates is merged into the existing workbook. Duplicate rows are removed automatically. Use this when adding new plates to an ongoing experiment.

**N - New file:** The existing workbook is kept and the new consolidation is saved under a different name. You will be prompted to enter a suffix, or press Enter to use a timestamp automatically.

## Understanding your output files

### Cleaning pipeline outputs (in the `outputs/` folder)

For each plate CSV processed, two CSV files are written:

**`[platename]_all_samples.csv`** - Every experimental row after cleaning. Standards, NTCs, and rows with no sample name have been removed, but all remaining rows are present regardless of whether they passed QC.

**`[platename]_review_samples.csv`** - Only rows flagged as needing manual review. This is a subset of the `_all_samples` file. If a replicate group is flagged, both rows of that pair appear here.

### Column descriptions

| Column | Source | Description |
|--------|--------|-------------|
| `Well` | CFX Maestro | Well position (e.g. A01, B02) |
| `Fluor` | CFX Maestro | Fluorophore channel (e.g. FAM, HEX) |
| `Target` | CFX Maestro | Gene/target name |
| `Content` | CFX Maestro | Well type from CFX Maestro (Unkn, Std-001, NTC, etc.) |
| `Sample` | CFX Maestro | Your sample identifier |
| `Cq` | CFX Maestro | Quantification cycle value. Blank for no-amplification samples or groups with >2 replicates. |
| `Starting Quantity (SQ)` | Modified by pipeline | SQ value. If the original was below LOD Lo or absent, set to LOD Lo ÷ 2. Blanked for groups with >2 replicates. |
| `DeltaCq` | Added by pipeline | Absolute difference between replicate Cq values. Only on the first row of each pair. |
| `AverageSQ` | Added by pipeline | Mean of replicate SQ values after LOD Lo adjustment. Only on the first row of each pair. |
| `review_reason` | Added by pipeline | Plain-English explanation of why a row was flagged. Blank if the row passed all checks. Multiple reasons separated by semicolons. |

### Review flags - what they mean and what to do

| Flag | What it means | Common causes | Suggested action |
|------|--------------|---------------|-----------------|
| `\|DeltaCq\| exceeds threshold` | Duplicate Cq values differ by more than 1.0 cycle. | Pipetting error, well-to-well variation, partial inhibition. | Check amplification curves in CFX Maestro. If one looks abnormal, exclude that replicate and note it. |
| `Only one replicate present` | Only a single well found for this sample/target. | Replicate not plated, failed to amplify, or excluded upstream. | Check the plate layout. Determine whether the single value is usable or if re-running is needed. |
| `One Cq is NA/NaN and the other is numeric` | One replicate amplified and one did not. | Partial inhibition, failed well, insufficient template. | Inspect both wells in CFX Maestro. Consider excluding the non-amplifying well. |
| `More than 2 replicates - manual reconciliation required` | Three or more wells found for this group. Cq and SQ are blanked. | Sample plated in triplicate, mislabelled layout, or two plates with the same sample names. | **Required action:** open `_all_samples.csv`, locate the group, determine the intended duplicates, and enter correct values manually. |
| `AverageSQ > LOD_Hi` | Average SQ exceeds the top of the standard curve. | Very high bacterial load, insufficient dilution. | Consider repeating with a higher dilution. Note the limitation if using the result. |
| `Unexpected negative for always-positive target (No amplification)` | Universal target produced no Cq value in any replicate. | Degraded DNA, very low concentration, failed extraction. | Investigate extraction quality before drawing conclusions from any target for this sample. |
| `Unexpected negative for always-positive target (Below LOD_Lo)` | Universal target produced Cq but SQ was below LOD Lo. | Very low bacterial DNA, poor extraction, borderline inhibition. | Interpret all results for this sample with caution. |
| `Sample name mismatch within (Fluor, Target, Content)` | Two or more different sample names found for the same Content/Target/Fluor combination. | Sample given different names in replicate wells, copy-paste error. | Check and correct sample names in CFX Maestro. Re-export and reprocess. |

### Consolidated Excel workbooks (in the `consolidated/` folder)

The consolidation script produces two Excel workbooks: `qpcr_all_samples.xlsx` and `qpcr_review_samples.xlsx`.

**Sheet layout:**

- The first sheet is always `run_summary` - a dashboard showing QC counts, standards check outcomes, and sample counts per plate and per target
- Each subsequent sheet contains data for one target (e.g. `fucp`, `hpd3`, `lyta`, `uni`)
- Each sheet is formatted as an Excel Table with auto-filters and a frozen header row

**Additional columns added by the consolidation script:**

| Column | Description |
|--------|-------------|
| `File Name` | The plate CSV filename the row came from. |
| `Date` | Run date parsed from the filename (YYYY-MM-DD). Blank if not parseable. |
| `Plate#` | Plate number parsed from the filename. Blank if not parseable. |
| `Repeat#` | Repeat number parsed from the filename. Blank if not present. |

**The `run_summary` sheet contains four sections:**

1. **Decision Breakdown** - how many times each QC rule fired per plate
2. **Standards Check Results** - pass/skipped/forced outcome per plate
3. **Final Sample Counts** - plate × target row counts for `all_samples`
4. **Review Sample Counts** - plate × target row counts for `review_samples`

## The audit folder

Every run writes to a set of log files in the `audit/` folder. These provide a complete, timestamped record of every decision made during processing.

### `pcr_decisions.csv`

One row per QC decision made on every sample and target.

| Column | Description |
|--------|-------------|
| `timestamp` | UTC date and time the decision was recorded |
| `user` | Windows/Mac username of whoever ran the script |
| `run_id` | Unique identifier for the plate processing run (format: `YYYYMMDD_HHMMSS_platename`) |
| `input_file` | The original plate CSV filename |
| `sample_id` | The sample identifier the decision applies to |
| `target` | The target (gene) the decision applies to |
| `rule_id` | The QC rule evaluated (e.g. `RV_DELTA_CQ`, `PASS_NEGATIVE`) |
| `outcome` | Whether the rule was `applied`, `skipped`, or resulted in a `pass` |
| `evidence` | Brief description of the values that led to the decision |
| `source` | Always `R_script` |
| `version` | Script version number |

> **Practical use:** To review all decisions for a specific plate, filter the `input_file` column. To see only flagged results, filter `outcome` to `applied`. To compare two runs of the same plate, filter by `run_id`.

### `pcr_variables.csv`

One row per configuration parameter used during each plate's processing run.

| Column | Description |
|--------|-------------|
| `var_name` | Parameter name (e.g. `dCq_thr`, `lod_hi`, `lod_lo`, `rm_patterns`) |
| `var_value` | The value of that parameter at the time of the run |
| `var_class` | The R data type of the value |

> **Practical use:** To verify what thresholds were applied to a historical plate run, filter by `run_id` and look at the `dCq_thr`, `lod_hi`, and `lod_lo` rows.

### `pcr_standards.csv`

One row per plate per run, recording the outcome of the standards check.

| Column | Description |
|--------|-------------|
| `n_expected` | How many standards were expected |
| `expected_standards` | Full list of expected labels, separated by pipes |
| `found_standards` | Standard labels actually found in the plate |
| `missing_standards` | Standards expected but not found. Empty if the plate passed. |
| `action` | `pass`, `skipped`, or `forced` |
| `lod_override` | LOD values entered manually if `action` was `forced`. Otherwise blank. |
| `notes` | Additional context |

### `pcr_run_log.txt`

A plain-text transcript of everything printed to the R console during the cleaning pipeline run. Overwritten each time the script runs.

> **Note:** If the script exits with an error before completing, the log file may be left in an incomplete state. If subsequent console output stops appearing in RStudio, type `sink()` in the console and press Enter to restore normal output.

### `pcr_consolidation_log.txt`

Equivalent to `pcr_run_log.txt` but for the consolidation script.

### `file_tree.txt`

A snapshot of the folder structure and file list discovered at the start of each cleaning pipeline run.

If something doesn't look right, see `vignette("troubleshooting", package = "qpcrpipeline")`.
