---
title: "Reference: Argument and Target Defaults"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reference: Argument and Target Defaults}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette lists the default values for every argument to `run_cleaning_pipeline()` and `run_consolidation_pipeline()`, and the default target/LOD lookup tables shipped with the package. For explanations and examples of how these are used, see `vignette("cleaning-pipeline", package = "qpcrpipeline")` and `vignette("consolidation", package = "qpcrpipeline")`, or `?run_cleaning_pipeline` / `?run_consolidation_pipeline`.

## `run_cleaning_pipeline()`

### Arguments

| Argument | Default |
|----------|---------|
| `input_dir` | `"RawData/"` |
| `file_pattern` | `"\\.csv$"` |
| `files` | `NULL` |
| `search_depth` | `2` |
| `tree_output` | `"both"` |
| `tree_path` | `"audit/file_tree.txt"` |
| `output_dir` | `"outputs"` |
| `dec_log_path` | `"audit/pcr_decisions.csv"` |
| `var_log_path` | `"audit/pcr_variables.csv"` |
| `delta_cq_threshold` | `1` |
| `rm_patterns` | `c("Std", "NTC", "Neg")` |
| `rm_columns` | `c("sample")` |
| `enable_preview` | `FALSE` |
| `dry_run` | `FALSE` |
| `debug_print` | `FALSE` |
| `skip_completed` | `TRUE` |
| `std_check_enabled` | `TRUE` |
| `n_standards` | `6` |
| `std_force_lod` | `NULL` |
| `std_log_path` | `"audit/pcr_standards.csv"` |
| `run_log_path` | `"audit/pcr_run_log.txt"` |
| `always_positive_targets` | `DEFAULT_always_positive_targets` |
| `target_lod` | `DEFAULT_target_lod` |

`always_positive_targets` and `target_lod` default to two exported objects, listed below, rather than literal values - pass your own list or vector to override them.

### `DEFAULT_always_positive_targets`

```{r}
DEFAULT_always_positive_targets
#> [1] "uni"       "univ"      "universal"
```

### `DEFAULT_target_lod`

`DEFAULT_target_lod` is a named list with two elements, `LOD_Hi` and `LOD_Lo`, each mapping target name spellings to their limit of detection:

| Target | LOD Hi | LOD Lo | Included alternative names |
|--------|--------|--------|------------------------------------|
| `fucp` | 2000 | 0.012 | hi_fucp, hi-fucp, hi fucp |
| `hpd3` | 2000 | 0.012 | hi_hpd3, hi-hpd3, hh_hpd3, hh_hypd |
| `lyta` | 2000 | 0.012 | |
| `copb` | 2000 | 0.012 | |
| `speb` | 2000 | 0.012 | |
| `nuc` | 2000 | 0.012 | |
| `gyrb` | 2000 | 0.012 | |
| `uni` | 200 | 0.0012 | univ, universal |

To add a new target, extend a copy of `DEFAULT_target_lod` and pass it via the `target_lod` argument:

```{r}
my_lod <- DEFAULT_target_lod
my_lod$LOD_Hi$newgene <- 2000
my_lod$LOD_Lo$newgene <- 0.012

run_cleaning_pipeline(target_lod = my_lod)
```

## `run_consolidation_pipeline()`

### Arguments

| Argument | Default |
|----------|---------|
| `input_dir` | `"outputs/"` |
| `all_pattern` | `"_all_samples\\.csv$"` |
| `review_pattern` | `"_review_samples\\.csv$"` |
| `consolidation_dir` | `"consolidated"` |
| `all_out_path` | `NULL` (resolves to `consolidated/qpcr_all_samples.xlsx`) |
| `review_out_path` | `NULL` (resolves to `consolidated/qpcr_review_samples.xlsx`) |
| `table_style` | `"TableStyleMedium2"` |
| `font_name` | `"Arial"` |
| `font_size` | `10` |
| `col_width_min` | `10` |
| `max_write_tries` | `5` |
| `wait_secs` | `3` |
| `target_aliases` | `list(uni = c("uni", "universal", "univ"))` |
| `update_target_to_canonical` | `TRUE` |
| `audit_dir` | `"audit/"` |
| `consolidation_log_path` | `"audit/pcr_consolidation_log.txt"` |

### Target aliases

| Canonical | Targets |
|-----------|---------|
| `uni` | uni, univ, universal |
