---
title: "Troubleshooting and Quick Reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting and Quick Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Troubleshooting

### Script will not start or package errors

**`Error: tidyverse is not installed`**

Run `install.packages("tidyverse")` in the R console and wait for it to complete, then try again.

**`Error: dplyr >= 1.1.0 required (installed: 1.0.5)`**

Your installed version is too old. Run `install.packages("tidyverse")` to update.

**Console output stops appearing after a script error**

Type the following in the console and press Enter:

```{r}
sink()
```

This closes the log file connection the script opened. If the problem persists, restart RStudio.

### File discovery problems

**`Error: No matching files found`**

Check:

- Your working directory in RStudio is set to the project folder (**Session -> Set Working Directory -> To Source File Location**)
- The `RawData` folder exists and contains `.csv` files
- The `input_dir` setting in Section 1 matches your actual folder name (case-sensitive on Mac and Linux)

**`Could not find a row with 'Well' in column 1`**

The file does not contain the expected `Well` header row. Possible causes:

- The file was saved from different software or a different export format
- The file was opened and re-saved in Excel, which may have corrupted the format - always work with copies, not originals
- Ensure you are exporting as "Quantification Results" CSV from CFX Maestro

**`Missing required columns`**

A required column is missing or has an unexpected name. Most common cause: the plate was exported without a required column, or a column was renamed after export. Always use the original, unmodified export file.

### LOD and target errors

**`Error: Missing targets in LOD_Hi: newgene`**

A target in your data is not defined in Section 2. Add the target's LOD values to Section 2 following the existing format.

**`Error: Different Hi values found: 200, 2000`**

Your plate contains targets with different LOD values (e.g. `uni` and `fucp` on the same plate file). Split the plate's CSV data into separate files - one per LOD group - before running the script.

### Standards check failures

**A plate repeatedly fails the standards check**

Check:

- Standards in the Content column must be labelled exactly as `Std-001` through `Std-006` (or `Std-01` through `Std-06`). Other formats such as `Standard 1`, `S1`, or `STD001` will not be recognised.
- If your assay uses a different number of standards, update `n_standards` in Section 1.
- If you run standards on a single plate and apply the curve to others, set `std_check_enabled <- FALSE` in Section 1 and provide LOD values via `std_force_lod`.

### Output file problems

**`[write retry 1/5] Could not write to '...'`**

The script retries up to 5 times. This usually means the output file is open in Excel. Close it and the script will succeed on the next retry.

**`Error: No matching files found in outputs/`**

Run the cleaning pipeline (Script 1) first, or check that `input_dir` in the consolidation script matches `output_dir` from the cleaning pipeline.

### Unexpected results in the output

**My sample rows are missing from the output**

Most common reasons:

- Sample name contained a reserved word (`Std`, `NTC`, or `Neg`) - rows are removed in Step 2
- Sample name was blank - rows are removed in Step 0
- Plate was skipped - check the run summary table in the console output

**DeltaCq and AverageSQ are blank for some rows**

Expected in two situations:

- **Second row of a replicate pair** - these values are only on the first row of each pair. Look one row up.
- **More than 2 replicates (`RV_EXCESS_REPS` flag)** - all Cq, SQ, DeltaCq, and AverageSQ are deliberately blanked. The `review_reason` column will explain this. You must resolve it manually.

**A target is not showing as a sheet in Excel**

Check:

- The target name in the CSV must match a name in Section 2 of the cleaning script
- Check `TARGET_ALIASES` in the consolidation script if the target uses multiple spellings
- Target names are normalised to lowercase - look for the sheet under the lowercase version of the name

## Quick reference

### Standard workflow

1. Export plate data from CFX Maestro as CSV (Quantification Results format, no structural modifications)
2. Rename each file to `YYMMDD_PlateN_[targets].csv` and place in the `RawData/` folder
3. Open `qpcr_cleaning_pipeline.R` in RStudio. Verify `input_dir` points to your `RawData` folder
4. Press **Ctrl+Shift+Enter** to run Script 1. Respond to any interactive prompts
5. Check the run summary - confirm all plates show `processed` status and row counts look reasonable
6. Open `qpcr_consolidation.R`. Verify `input_dir` is set to `outputs/`
7. Press **Ctrl+Shift+Enter** to run Script 2. Choose O/A/N if prompted about existing workbooks
8. Open `qpcr_all_samples.xlsx` and `qpcr_review_samples.xlsx` from the `consolidated/` folder
9. Review any rows in the review workbook and investigate flagged results before using them in your analysis

### Reserved words - never use in sample names

| Reserved word | What it matches (examples) | Effect |
|--------------|---------------------------|--------|
| `Std` | Std, Std-001, STD_curve, MyStd | Row removed from output |
| `NTC` | NTC, ntc, NTC_plate2 | Row removed from output |
| `Neg` | Neg, NegControl, Negative_sample, neg | Row removed from output |

### Review flag summary

| `review_reason` text | Meaning | Urgency |
|----------------------|---------|---------|
| `\|DeltaCq\| exceeds threshold` | Duplicates differ by > 1 Cq | Review |
| `Only one replicate present` | Single well only | Review |
| `One Cq is NA/NaN and the other is numeric` | Mixed amplification in duplicates | Review |
| `More than 2 replicates - manual reconciliation required` | 3+ wells for this group; Cq/SQ blanked | **Action required** |
| `AverageSQ > LOD_Hi` | Above standard curve range | Review |
| `Unexpected negative for always-positive target (No amplification)` | uni/univ/universal did not amplify | Investigate sample quality |
| `Unexpected negative for always-positive target (Below LOD_Lo)` | uni/univ/universal below detection | Investigate sample quality |
| `Sample name mismatch within (Fluor, Target, Content)` | Inconsistent sample labelling | Correct in CFX Maestro and reprocess |

### Interactive prompts at a glance

| Prompt | Options | Choose when |
|--------|---------|-------------|
| Skip completed plates? | Y = skip, N = reprocess all | Y when adding new plates to an existing experiment. N when you have changed data or settings. |
| Standards check failed | S = skip, M = mixed handling, F = force all | S if plate data is unreliable. M if only interior standards are missing. F if you have manually verified LOD values. |
| Blank sample names | S = skip, C = use Content as sample name | S if you want to add sample names first. C only as a temporary measure. |
| Output workbook already exists | O = overwrite, A = append, N = new file | O for a fresh run. A when adding new plates. N to preserve the existing version. |
